---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# r_clustering

The goal of r_clustering is to demonstrate a simple r package, doc, and testing infrastructure for DS purposes.

## Installation

You cannot install r_clustering from [CRAN](https://CRAN.R-project.org) because the package name is not valid.


You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("Henrik-Kowalkowski/r_clustering")
```
## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(r_clustering)
clusts <- make_clusters(4, c_range = 20, density = 500, categorical = F, seed = 1)
```

Here's how to visualize the clusters:

```{r cars}
library(dplyr)
library(tidyselect)
library(tidyr)
library(stringr)
library(ggplot2)
library(gridExtra)

clusts <- make_clusters(4, c_range = 20, density = 500, categorical = F, seed = 1)

clust_positions <- clusts %>%
  group_by(y_true) %>%
  summarize_all(list(min = min, max = max, mean = mean)) %>%
  select(sort(peek_vars()))

clust_plot_data <- clust_positions %>%
  pivot_longer(cols = !y_true) %>%
  mutate(coord = if_else(str_detect(name, "^y_"), "y", "x")) %>%
  mutate(name = str_sub(name, start = 3)) %>%
  pivot_wider(names_from = coord, values_from = value) %>%
  mutate(center = factor(name == "mean"))

shapes <- c(16, 3)
names(shapes) <- c(F, T)
sizes <- c(3, 10)
names(sizes) <- c(F, T)


p1 <- clust_plot_data %>% ggplot(aes(x = x, y = y, col = y_true, shape = center, size=center)) +
  geom_point() +
  coord_cartesian(xlim = c(-20, 20), ylim = c(-20, 20)) +
  scale_shape_manual(values = shapes) + scale_size_manual(values = sizes) +
  theme(legend.position = "none")

p2 <- clusts %>% ggplot(aes(x=x, y=y, col=y_true)) + geom_point() + coord_cartesian(xlim = c(-20, 20), ylim = c(-20, 20))

grid.arrange(p1, p2, ncol=2)
```
